Confronting preferential sampling in count and occupancy surveys: diagnosis and model-based triage \footnotemark[2]

### 2015-07-22 manuscript/pref_sampling.tex
Bias correction for TMB now implemented correctly for random effects, but simple spatial regression models still point to a bias in posterior predictions (likely associated with fixed effect).  Plan now is to switch to MCMC via Gibbs sampling.
Updated MS to make model stucture for preferential sampling similar to Royle and Berliner (also now better reflects model in Pati et al.).  This will hopefully resolve some identifiability issues I ran into in TMB simulations.

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling preferential sampling.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the expected value of the spatial process model.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, if sampling intensity is high in areas close to bases of operations (e.g., landing strips in the case of aerial surveys), we might expect there to be a greater propensity for PS close to these bases (at least, if animal density is anomolously high or low in these areas).

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the process model, we start with Eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then expand the link-scale expectation from this model
(i.e. Eq. \ref{eq:mu}) as follows:
\begin{eqnarray}
  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\eta_i+\delta_i+\epsilon_i,
\label{eq:mu_aug}
\end{eqnarray}
where $\delta_i$ is an additional, spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
\begin{eqnarray}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
\begin{eqnarray}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\gamma_i+\delta_i+b \mu_i + \varepsilon_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\gamma_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models.  Following \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we also include the potential for a linear effect of $\mu_i$ on sampling intensity in Eq. \ref{eq:nu} (with parameter $b$).  However, including both $b \mu_i$ and $\delta_i$ in Eqs. \ref{eq:mu_aug} and \ref{eq:nu} will likely lead to an overparameterized model.  Useful constraints to place on these models are (a) $b=0$ (i.e., spatially varying PS), (b) $\delta_i = 0$ (i.e., a linear effect of the link-scale state process on sampling intensity), and (c) $b=\delta_i=0$ (i.e., no preferential sampling).  A comparison of the performance of models with different sets of constraints can serve as a test of PS.  In particular, if one can demonstrate that models with constraint (c) perform similarly or better than models with (a) and (b) constraints, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e. not modeling sampling intensity).

>>>>

The models considered by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011} are a useful first step in addressing and modeling preferential sampling.  However, they are somewhat limited since they are specific to continuous spatial domains, continuous data (as opposed to presence/absence or count data), and Gaussian error distributions.  Also, they require the linear predictor of the preferential sampling model to be written as a simple linear function of the the spatial process model for density.  In real world applications, we can envision cases where sampling is strongly preferential in certain areas of the landscape, and not in others.  For instance, sampling may be more strongly preferential close to bases of operations, (e.g., landing strips in the case of aerial surveys), but less so in areas that are harder to get to.

Given these limitations, our present task is to generalize PS models to the types of data more typical of SDMs, and to allow the degree of PS to vary across the landscape.  Like \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}, we impose a joint model for the process of interest (animal abundance or occurence) and the locations chosen for sampling. For the process model, we start with Eq. \ref{eq:process} as a general formulation for non-Gaussian data.  We then expand the link-scale expectation from this model
(i.e. Eq. \ref{eq:mu}) as follows:
\begin{eqnarray}
  \mu_i & = & \beta_0 + {\bf x}_i \boldsymbol{\beta}+\delta_i+\epsilon_i,
\label{eq:mu_aug}
\end{eqnarray}
where $\delta_i$ is a spatially referenced random effect.  Next, we model the binary indicator for sample inclusion using a Bernoulli distribution:
\begin{eqnarray}
  R_i & \sim & \text{Bernoulli}(h^{-1}(\nu_i)),
\end{eqnarray}
where $h()$ denotes a link function appropriate for binary data (e.g. logit, probit).  We then write the linear predictor for this model as
\begin{eqnarray}
  \nu_i & = & \beta_0^* + {\bf x}_i^* \boldsymbol{\beta}^*+\eta_i+{\bf B} \delta_i + \varepsilon_i.
\label{eq:nu}
\end{eqnarray}
 In a similar fashion to the model for the state process, the sampling intensity model has an intercept ($\beta_0^*$), explanatory covariates (${\bf x}_i^*$), fixed effect regression parameters ($\boldsymbol{\beta}^*$), spatially autocorrelated random effects ($\eta_i$ and $\delta_i$), and normally distributed error $\varepsilon_i$.  The predictive covariates ${\bf x}_i$ from Eq. \ref{eq:mu_aug} and ${\bf x}_i^*$ from Eq. \ref{eq:nu} need not be the same (although they can be).  Note also that the  spatially autocorrelated random effect $\delta_i$ is included in both Eqs. \ref{eq:mu_aug} and \ref{eq:nu}, allowing for dependency in the two models, with the matrix {\bf B} describing the strength and type of dependence between the sampling process and underlying density.

 The formulation in Eq. \ref{eq:nu} is the same as previously proposed by other authors for hierarchical multivariate models with spatial dependence \citep[cf.][]{RoyleBerliner1999}.  There are multiple ways of structuring ${\bf B}$ depending on the complexity with which one wants spatial dependence that is desired \citep{RoyleBerliner1999}.  For instance, setting ${\bf B} = {\bf O}_{S \times S}$ corresponds to an absence of spatial dependence (and thus no preferential sampling).  Setting ${\bf B} = b {\bf I}$, where $b$ is an estimated parameter and {\bf I} is an $(S \times S)$ identity matrix corresponds to the linear preferential sampling model suggested by \citet{DiggleEtAl2010} and \citet{PatiEtAl2011}.  Alternatively, we could allow the degree of PS to vary across the landscape.  For instance, one can contemplate a trend surface model for preferential sampling by specifying a diagonal matrix for ${\bf B}$, with entries given by $b_0 + b_1 \textrm{lat}_i + b_2 \textrm{long}_i$, where $b_0$, $b_1$, and $b_2$ are estimated parameters and $\textrm{lat}_i$ and $\textrm{long}_i$ give latitude and longitude, respectively \citep{RoyleBerliner1999}.  Theoretically, one could include more highly parameterized structures for spatial dependence, such as higher order trend surfaces or a two dimensional spline \citep{RoyleBerliner1999,UnpublishedManuscript}, but the ability to robustly estimate the parameters of such a model is likely dependent on having a rich, spatially balanced dataset, which is often not the case in ecological applications.

A comparison of the performance of models with different sets of constraints on ${\bf B}$ can serve as a test of PS.  In particular, if one can demonstrate that models with ${\bf B}={\bf 0}$ perform similarly or better than models with ${\bf B} \ne {\bf 0}$, then PS is likely not worth modeling and inference can proceed using standard SDMs (i.e. not modeling sampling intensity).
